HealpixNet: Efficient spherical Convolutional Neural Network with Healpix sampling for cosmological applications

### 2018-07-30 scnn.tex
working on section 4

\newcommand{\nati}[1]{{\color[rgb]{.1,.6,.1}{#1}}}
\newcommand{\TK}[1]{{\color{red}{#1}}}
\newcommand{\todo}[1]{{\color[rgb]{.6,.1,.6}{#1}}}
\newcommand{\assign}[1]{{\color[rgb]{.8,.5,.8}{Assigned: #1 }}}

@@

\citet{boomsma2017spherical} apply a standard 2D CNN to a grid discretization of the sphere. Two grids, illustrated in \figref{sphere_grids}, are used in their work: the equirectangular grid, and the cubed-sphere grid defined by~\citet{ronchi1996cubed}.
As this formulation uses the standard 2D convolutions, all the optimizations developed for 2D images can be applied, what make it computationally very efficient.
It is, however, restricted to those grid sampling sets. Similarly to the cubed-sphere, HEALPix is a composition of twelve grids. While each of those twelve grids could potentially be independently fed into a 2D CNN, boundary conditions would have to be properly handled, as those grids are not independent.
\todo{How do thy handle boundaries?}
\TK{+1,good question}
% Furthermore, the grid requirement makes it impossible for the convolution to capture the spherical properties of the domain, i.e., the cubed-sphere sampling is by definition adapted to a cube and not to a sphere. Nope, the graph can adapt to any sampling.
% Moreover, those samplings are not ideal as their resolution is not constant across the sphere.
Moreover, the pixel area is not constant across the sphere in those sampling schemes, which may limit their potential applications.
Indeed, the singularities, such as the poles for the equirectangular grid, and the corners of the cube for the cubed-sphere, are over-sampled.
Equal areas of discrete elements is advantageous for many physical measurements, as white noise at the sampling frequency of the instrument gets integrated exactly into white noise in the pixel space, and sky signals are sampled without regional dependence \citep{gorski2005healpix}.
% \nati{Michael: Under some very specific hypothesis, this second approach is a particular case of our method, i.e. (i) our method with a stupid sampling, (ii) assumption that graph convolution on a grid == 2d convolution. Do you think we should mention that? I think we should not.}
Finally, a 2D CNN learns directional filters which orientation is arbitrary. Indeed, directions, which make much sense for images (up, down, left, and right), are arbitrarily defined when setting the origin of the grids on the sphere. Any convolution using directions on the sphere is not equivariant to translation and rotation.

@@

\citet{cohen2018spherical} uses the continuous spherical convolution.
The goal of this contribution was to define a rotation-equivariant convolution on the sphere, i.e., a rotation of the input implies the same rotation in the output.
The resulting convolution is performed by (i) a spherical harmonic transform (SHT), i.e., a projection on the spherical harmonics, (ii) a  multiplication in the spectral domain, and (iii) an inverse SHT. Hence the computational cost of a convolution is essentially due to the two SHTs. A naive implementation of the SHT costs $\bO(N_{pix}^2)$ operations, where $N_{pix}$ is the number of pixels. There fortunately exists accelerated schemes for some sampling sets, see for example \citet{mohlenkamp1999fast, rokhlin2006fast, reinecke2013libsharp}.
The main advantage of this approach is that it provides a mathematically well defined rotation-equivariant network. Nevertheless, even with accelerated SHTs, the convolutions remain expensive, limiting the practical use of this approach.
For example with the HEALPix sampling, which was designed to have a fast SHT by being iso-latitude, the computational cost of the SHT is $\bO(N_{pix}^{3/2}) = \bO(N_{side}^3) = \bO(\ell_{max}^3)$, where $\ell_{max}$ is the largest order (angular frequency) \citep{gorski2005healpix, reinecke2013libsharp}.\footnote{All pixels are placed on $N_{ring} = 4N_{side}-1 = \bO(\sqrt{N_{pix}})$ rings of constant latitude. Each ring has $\bO(\sqrt{N_{pix}})$ pixels. Thanks to this iso-latitude property, the SHT is computed using recurrence relations for Legendre polynomials on co-latitude and fast Fourier transforms (FFTs) on longitude. The computational cost is thus $\sqrt{N_{pix}}$ FFTs for a total cost of $\bO(N_{pix} \log(\sqrt{N_{pix}}))$, plus $\sqrt{N_{pix}}$ matrix multiplications of size $\sqrt{N_{pix}}$ for a total cost of $\bO(N_{pix}^{3/2})$.}
In comparison, filtering with our method scales as $\bO(N_{pix})$ (see \secref{efficient_convolution}). Moreover, \figref{filtering_speed} shows that a naive implementation of our method is more than ten times faster than the SHT implemented by the highly optimized \pkg{libsharp} library used by HEALPix \citep{reinecke2013libsharp}.
While a comparison is theoretically possible, the experiments of~\cite{cohen2018spherical} are done using the geodesic grid sampling set and do not provide any code for the HEALPix one. Hence, a practical comparison is not possible.
We also note that this approach cannot be easily accelerated when data lies on a part of the sphere only. That is an important use case in cosmology as measurements are often partial, i.e., full-sky maps are rare. One could still fill the unseen part of the sphere with zeros, and avoid to compute empty integrals (on the latitude or longitude). It is however not straightforward to identify those empty directions, and computations would still be wasted (consider for example a ring which mostly contains zeros but a few measurements). With graphs however, one only pays for the used pixels. The cost is some distortions due to border effects (see \figref{border_effects} and \ref{sec:border_effects}). Those can however be mitigated by padding with zeros a small area around the measurements.

In an attempt to get the best of the first two approaches, we follow a third
direction. We use a graph to adapt to the particular structure of the sphere.
While the convolution remains efficient, it still captures well the spherical
structure and is particularly adapted to the HEALPix sampling.

The use of a graph to model the discretized sphere was considered by~\citet{khasanova2017graphomni} as well. The authors map omnidirectional images to the sphere, with the goal of correcting for distortions induced by the lens used for the acquisition.
Besides the different application covered by the paper, there are two important
differences. First the sampling is different as~\cite{khasanova2017graphomni} uses a
grid. Second, it uses a different graph CNN based on a different parameterization
of the convolution kernel.
As~\citep{boomsma2017spherical}, they use an equirectangular grid sampling. Moreover, they did not take advantage of a hierarchical pixelization of the sphere and resorted to dynamic pooling~\citep{kalchbrenner2014dcnn}.
Finally, while we do want our network to be rotation invariant, their statistical layer --- an operation which computes a set of spatially invariant statistics of the last feature map --- destroys a lot of information \todo{and hinders performance (should be empirically verified)}. Moreover, the chosen statistics have to be hand-crafted to capture statistics of interest. While developing completely invariant transforms is a  goal worth investigating, CNNs for image recognition (arguably the most researched and developed architectures), are not built to be fully invariant to translation. As with the proposed spherical CNN, the fully connected layers at the end of the network depend on the spatial localization of features. While that may not be ideal, it was shown to work well in practice.
\todo{In this work we reused the idea of a statistical layer, described in \secref{stat_layer}, to provide invariance to isometric transformations such as translation and rotation on the sphere. While \citet{khasanova2017graphomni} uses the mean and variance of diffused versions of the last feature maps, we experimentally tested multiple variants and got good results by keeping all the statistical moments via a learned histogram. That is similar to the learnable histogram introduced by \citet{wang2016learnhist}, although for the orthogonal purpose of integrating global context in object detection and segmentation tasks.}

@@

In this work, we are using the graph CNN formulation introduced in~\cite{defferrard2016convolutional}.

@@

A CNN is composed of the following main building blocks~\citep{lecun1998cnn}: (i) a convolution, (ii) a non-linearity, (iii) a down-sampling operation, (iv) a pooling operation, and (v), optionally, some normalization.\footnote{\citet{ioffe2015batchnorm} have shown that batch normalization helps training. We verified this experimentally in our setting as well.} Our architecture is depicted in~\figref{architecture} and discussed in greater details in \secref{architecture}. As operations (ii) and (v) are point-wise, they do not depend on the data domain. The pooling operation is simply a permutation invariant aggregation function which does not need to be adapted either. The convolution and down-sampling operations however need to be generalized from Euclidean domains to the sphere.

On regular Euclidean domains, such as 1-dimensional time series or 2-dimensional images, a convolution can be efficiently implemented by sliding a localized convolution kernel (for example a patch of 5x5 pixels) in the signal domain. Likewise, down-sampling can be achieved by taking one pixel out of $n$.
Because of the irregular sampling, there is no straightforward way to define a convolution on the sphere directly in the pixel domain.

We will hence make a detour by the spectrum to perform the convolution.\footnote{Note that we only use the spectrum to define the convolution. The implementation however does not need a direct access to the spectrum, which is computationally much more efficient. See \secref{efficient_convolution}.}
The gist of our method is to define the convolution on a sphere using a graph, and the down-sampling using a hierarchical pixelization of the sphere.

@@

Before doing any numerical analysis on the sphere, one first has to choose a tessellation, i.e., an exhaustive partition of the sphere into finite area elements, where the data under study is quantized.
The simplicity of the spherical form belies the intricacy of global analysis on the sphere. There is no known point set which achieves the analogue of uniform sampling in Euclidean space and allows exact and invertible discrete spherical harmonic decompositions of arbitrary but band-limited functions. Any existing proposition of practical schemes for the discrete treatment of such functions on the sphere introduces some systematic error dependent on the global properties of the point set.
While our method is applicable to any pixelization of the sphere, two details depend on the chosen sampling: (i) the choice of neighbors in the construction of the graph, and (ii) the choice of parent nodes when coarsening the graph.
As HEALPix~\citep{gorski2005healpix} is the most popular sampling used in cosmology, our target application, we'll tailor the method to that particular sampling in the subsequent exposition.
%We'll note in later sections where the method should be adapted for a different sampling.

@@

HEALPix is an acronym for Hierarchical Equal Area isoLatitude Pixelization of a sphere. As suggested in the name, this pixelization produces a hierarchical subdivision of a spherical surface in which each pixel covers the same surface area as every other pixel.
A hierarchy is desired for data locality in computer memory. Equal area is advantageous because white noise generated by the signal receiver gets integrated exactly into white noise in the pixel space. Isolatitude is essential for the implementation of a fast spherical transform. HEALPix is the sole pixelization scheme which satisfies those three properties.

The lowest possible resolution is given by the base partitioning of the surface into $N_{pix} = 12$ equal sized pixels (right-most sphere in \figref{pooling}). As each pixel is subdivided in four, the second coarser resolution is $N_{pix} = N_{side}^2 \cdot 12 = 2^2 \cdot 12 = 48$ pixels (middle sphere in \figref{pooling}), the third is $N_{pix} = N_{side}^2 \cdot 12 = 4^2 \cdot 12 = 192$ pixels, etc., where $N_{side} = 1, 2, 4, 8, \ldots$ is the grid resolution parameter.
High-resolutions HEALPix maps easily reach millions of pixels. \figref{example_maps} shows two examples.

@@

From the HEALPix pixelization, we build a weighted undirected graph $\G = (\V, \E, \W)$, where $\V$ is the set of $N_{pix} = |\V|$ nodes, $\E$ is the set of edges, and $\W$ is the weighted adjacency matrix. In our graph, each pixel $i$ is represented by a node (also called vertex) $v_i \in \V$. Each node $v_i$ is then connected to the $8$ (or $7$)\footnote{\label{neighbors}The $12 \cdot 4 = 48$ pixels at the corner of each rhombus of the base dodecahedron only have 7 neighboring pixels. See Figures~\ref{fig:healpix_graph_4} and~\ref{fig:pooling}.} nodes $v_j$ which represent the neighboring pixels $j$ of pixel $i$, forming edges $(v_i, v_j) \in \E$. Given those edges, we define the weighted adjacency matrix $\W \in \R^{N_{pix} \times N_{pix}}$ as
\begin{equation*}
	\W_{ij} = \begin{cases}
		\exp \left( -\frac{\|\x_i-\x_j\|_2^2}{\sigma^2} \right) & \text{if pixels $i$ and $j$ are neighbors,} \\
		0 & \text{otherwise,} \\
	\end{cases}
\end{equation*}
where $\x_i$ is a vector encoding the 3-dimensional coordinates of pixel $i$, and
\begin{equation*}
	\sigma = \frac{1}{|\E|} \sum_{(v_i, v_j) \in \E} \|\x_i-\x_j\|_2
\end{equation*}
is the average Euclidean distance over all connected pixels. This weighting scheme is important as while pixel areas are equal, distances between pixels are not.
Other weighting schemes are possible. For example,~\cite{khasanova2017graphomni} uses the inverse of the distance instead. We found out that the one proposed above works fine for our purpose.
\figref{healpix_graph_4} shows a graph constructed from the HEALPix sampling of a sphere.

@@

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{eigenvectors}
	\caption{The first 16 Fourier modes (eigenvectors of the graph Laplacian) of a graph constructed from the HEALPix sampling of the sphere. Eigenvectors 1--3 could be associated with spherical harmonics of degree $\ell=1$ and order $|m|=(0,1)$, eigenvectors 4--8 with degree $\ell=2$ and order $|m|=(0,1,2)$, and eigenvectors 9--15 with degree $\ell=3$ and order $|m|=(0,1,2,3)$. Nevertheless, the graph eigenvectors are \emph{only approximating} the real spherical harmonics.}
	\label{fig:graph_harmonics}
\end{figure}

It turns out that using the graph construction described above, the graph Fourier modes resemble the spherical harmonics (see \figref{graph_harmonics}). That is a strong hint that the graph is able to capture the spherical properties of the HEALPix sampling. This topic is further discussed in \ref{sec:comparison_spherical_harmonics}.

@@

As there is no notion of translation on a graph, we cannot really convolve two graph signals. We can, however, convolve a signal with a kernel defined in the spectral domain. More precisely, we are filtering a graph signal by a kernel. Given the convolution kernel
$h: \R_+ \rightarrow \R$, a signal $\f \in \R^{N_{pix}}$ on the graph is filtered as
\begin{equation} \label{eqn:graph_convolution}
	h(\L) \f = \U h(\bLambda) \U\trans \f,
\end{equation}
where $h(\bLambda)$ is a diagonal matrix where $h(\bLambda)_{ii} = h(\bLambda_{ii})$, and $h(\bLambda) \hat{\f}$ is the spectral representation of $\f$ after filtering.

Contrary to classical signal processing on Euclidean domains, the kernel $h$ has no single representation in the vertex domain and cannot be translated on the graph. It can however be localized on any vertex $v_i$ by the convolution with a Kronecker delta $\b \delta_i \in \R^{N_{pix}}$,\footnote{A Kronecker delta is the signal $\b \delta_i \in \R^{N_{pix}}$ that is zero everywhere except on vertex $v_i$ where it takes the value one.} that is $h(\L) \b \delta_i$, a column of $h(\L)$.
This localization can be useful to visualize kernels as explained in~\ref{sec:filter_visualization}.
If the graph is not regular, i.e., all nodes don't have the same number of neighbors, and all distances are not equal, the effect of the kernel will slightly differ from one node to another. While there is no perfect sampling of the sphere, these differences are negligible as the structure of the whole graph is very regular. However, when considering only parts of the sphere, one can observe important border effects (see \figref{border_effects} and~\ref{sec:border_effects}).
Finally, the graph convolution can be interpreted in the vertex domain as a scalar product with all localizations of the kernel:
\begin{equation} \label{eqn:graph_convolution_spatial}
	(h(\L) \f)_i = \langle h(\L) \b \delta_i, \f \rangle.
\end{equation}
\textbf{Example.}
Let us consider the heat diffusion problem
\begin{equation} \label{eqn:heat_equation}
	\L \b{f}(t) = \tau \partial_t \b{f}(t),
\end{equation}
where $\b{f}: \R_+ \rightarrow \R^{N_{pix}}$. Given the initial condition
$\b{f}(0)$, the solution of~\eqnref{heat_equation} can be expressed as
\begin{equation*}
	\b{f}(t) = e^{-\L \tau t} \b{f}(0) = \U e^{-\bLambda t \tau} \U\trans \g{f}(0) = K_t(\L) \b{f}(0),
\end{equation*}
which is, by definition, the convolution of the signal $\b{f}(0)$ with the kernel $K_t(\lambda)=e^{-\tau t \lambda}$. Since the kernel $K_t$ is applied to the graph eigenvalues $\bLambda$, which can be interpreted as squared frequencies, it can also be considered as a generalization of the Gaussian kernel on the sphere. \figref{gaussian_filters_comparizon} shows the effect of the convolution by diffusing a unit of heat for $\tau=1$ at various times $t$. By comparing the graph convolution with the spherical symmetric Gaussian smoothing, we observe that both techniques lead to similar results (see \figref{gaussian_filters_comparizon}). While the graph convolution remains different from the spherical convolution, this small experiment shows that, providing the correct parameters, graph convolution can approximate spherical convolution. For additional insides about the graph kernel $K_t$, we refer to the \figref{gaussian_filters_visualization} and~\ref{sec:filter_visualization}.

@@

Hierarchical pixelization schemes present a natural coarsening. Indeed, each subdivision divides a cell in an equal number of child sub-cells. Coarsening is the reverse operation: merging the sub-cells toward the goal of summarizing the data supported on them. Merging cells lead to a coarser graph, hence the name coarsening. Coarsening defines $\C(i)$, the set of children of node $v_i$. For the HEALPix subdivision scheme, the number of children is constant, i.e., $| \C(i) | = 4^p \ \forall i$, for some $p$.

>>>>

\newcommand{\nati}[1]{{\color[rgb]{.1,.6,.1}{#1}}}
\newcommand{\TK}[1]{{\color{red}{TK:#1}}}
\newcommand{\todo}[1]{{\color[rgb]{.6,.1,.6}{#1}}}
\newcommand{\assign}[1]{{\color[rgb]{.8,.5,.8}{Assigned: #1 }}}

@@

\citet{boomsma2017spherical} apply a standard 2D CNN to a grid discretization of the sphere. Two grids, illustrated in \figref{sphere_grids}, are used in their work: the equirectangular grid, and the cubed-sphere grid defined by~\citet{ronchi1996cubed}.
As this formulation uses the standard 2D convolutions, all the optimizations developed for 2D images can be applied, what make it computationally very efficient.
It is, however, restricted to those grid sampling sets. Similarly to the cubed-sphere, HEALPix is a composition of twelve grids. While each of those twelve grids could potentially be independently fed into a 2D CNN, boundary conditions would have to be properly handled, as those grids are not independent.
\todo{How do thy handle boundaries?}
\TK{+1,good question}
% Furthermore, the grid requirement makes it impossible for the convolution to capture the spherical properties of the domain, i.e., the cubed-sphere sampling is by definition adapted to a cube and not to a sphere. Nope, the graph can adapt to any sampling.
% Moreover, those samplings are not ideal as their resolution is not constant across the sphere.
Moreover, the pixel area is not constant across the sphere in those sampling schemes, which may limit their potential applications.
Indeed, the singularities, such as the poles for the equirectangular grid, and the corners of the cube for the cubed-sphere, are over-sampled.
Equal areas of discrete elements is advantageous for many physical measurements, as white noise at the sampling frequency of the instrument gets integrated exactly into white noise in the pixel space, and sky signals are sampled without regional dependence \citep{gorski2005healpix}.
% \nati{Michael: Under some very specific hypothesis, this second approach is a particular case of our method, i.e. (i) our method with a stupid sampling, (ii) assumption that graph convolution on a grid == 2d convolution. Do you think we should mention that? I think we should not.}
Finally, a 2D CNN learns directional filters, which are well defined with respect of the image coordinates.
Directions that make much sense for images (up, down, left, and right) are arbitrarily defined when setting the origin of the grids on the sphere.
Any convolution using directions on the sphere is not equivariant to translation and rotation.

@@

\citet{cohen2018spherical} uses the continuous spherical convolution.
The goal of this contribution was to define a rotation-equivariant convolution on the sphere, i.e., a rotation of the input implies the same rotation in the output.
The resulting convolution is performed by (i) a spherical harmonic transform (SHT), i.e., a projection on the spherical harmonics, (ii) a  multiplication in the spectral domain, and (iii) an inverse SHT. Hence the computational cost of a convolution is dominated by the two SHTs.
A naive implementation of the SHT costs $\bO(N_{pix}^2)$ operations, where $N_{pix}$ is the number of pixels.
Accelerated schemes exist for some sampling sets \citep[see][for examples]{mohlenkamp1999fast, rokhlin2006fast, reinecke2013libsharp}.
The main advantage of this approach is that it provides a mathematically well defined rotation-equivariant network. Nevertheless, even with accelerated SHTs, the convolutions remain expensive, limiting the practical use of this approach.
For example with the HEALPix sampling, which was designed to have a fast SHT by being iso-latitude, the computational cost of the SHT is $\bO(N_{pix}^{3/2}) = \bO(N_{side}^3) = \bO(\ell_{max}^3)$, where $\ell_{max}$ is the largest angular frequency\footnote{All pixels are placed on $N_{ring} = 4N_{side}-1 = \bO(\sqrt{N_{pix}})$ rings of constant latitude. Each ring has $\bO(\sqrt{N_{pix}})$ pixels. Thanks to this iso-latitude property, the SHT is computed using recurrence relations for Legendre polynomials on co-latitude and fast Fourier transforms (FFTs) on longitude. The computational cost is thus $\sqrt{N_{pix}}$ FFTs for a total cost of $\bO(N_{pix} \log(\sqrt{N_{pix}}))$, plus $\sqrt{N_{pix}}$ matrix multiplications of size $\sqrt{N_{pix}}$ for a total cost of $\bO(N_{pix}^{3/2})$.}
\citep{gorski2005healpix, reinecke2013libsharp}.
In comparison, filtering with our HealpixNet method scales as $\bO(N_{pix})$ (see \secref{efficient_convolution}).
% Moreover, \figref{filtering_speed} shows that a naive implementation of our method is more than ten times faster than the SHT implemented by the highly optimized \pkg{libsharp} library used by HEALPix \citep{reinecke2013libsharp}.
% While a comparison is theoretically possible, the experiments of~\cite{cohen2018spherical} are done using the geodesic grid sampling set and do not provide any code for the HEALPix one. Hence, a practical comparison is not possible.
% \TK{This I would leave out here, as this is a "spoiler" of the results section.}

Approaches by \citet{cohen2018spherical, boomsma2017spherical} can not be easily accelerated when the data lies on a part of the sphere only.
That is an important use case in cosmology as measurements are often partial, i.e., full-sky maps are rare.
One could still fill the unseen part of the sphere with zeros, and avoid to compute empty integrals, on the latitude or longitude.
It is, however, not straightforward to identify those empty directions, and computations would still be wasted; consider, for example, a ring which mostly contains zeros but a few measurements.
With graphs however, the computation is done only for the used pixels.
This, however, results in some distortions due to border effects (see \figref{border_effects} and \ref{sec:border_effects}).
These can be mitigated by padding with zeros a small area around the measurements.

% In an attempt to get the best of the first two approaches, we follow a third
% direction. We use a graph to adapt to the particular structure of the sphere.
% While the convolution remains efficient, it still captures well the spherical
% structure and is particularly adapted to the HEALPix sampling.
% \TK{I am not sure what is the point of this paragraph, we described our approach already before}

The use of a graph to model the discretised sphere was also considered by~\citet{khasanova2017graphomni}.
The authors map omnidirectional images onto a sphere, with the goal of correcting for distortions induced by the lens, which was used for their acquisition.
Besides a different practical application, for which the method was designed, there are two important differences.
% First, a different sampling scheme is used: \cite{khasanova2017graphomni} uses a grid.
First, similarly to~\citet{boomsma2017spherical}, they use an equirectangular grid sampling.
% \TK{how is it different from us?}
Second, it uses a graph CNN based on a different parameterisation of the convolution kernel.
Moreover, they did not take advantage of a hierarchical pixelization of the sphere and resorted to dynamic pooling~\citep{kalchbrenner2014dcnn}.
Finally, while the rotation invariance is desirable, the statistical layer used
% --- an operation which computes a set of spatially invariant statistics of the last feature map ---
% \{TK} this should be explained earlier
may not capture the information in an optimal way
\todo{and hinders performance (should be empirically verified)}.
\TK{This is a strong claim, can we support this?}
Moreover, the chosen statistics have to be hand-crafted to capture statistics of interest.
% While developing completely invariant transforms is a goal worth investigating, CNNs for image recognition, which are arguably the most researched and developed architectures, are not built to be fully invariant to translation.
% As with the proposed spherical CNN, the fully connected layers at the end of the network depend on the spatial localization of features.
% While that may not be ideal, it was shown to work well in practice.
% \TK{The two sentences above look like a tangent to me.. can we skip it an go to the main point, which is I think the fact that we use the stat layer?}

\todo{In HealpixNet we use the idea of a statistical layer, described in \secref{stat_layer}, to provide invariance to isometric transformations such as translation and rotation on the sphere.
\TK{We would need to describe the idea of the statistical layer for the cosmology audience. They don't have a clue what it is.}
The use of the statistical layer is optional.
In our approach, we are keeping all the statistical moments by retaining a full, learned learned feature histogram, an approach that performed the best in our tests.
That is similar to the learnable histogram introduced by \citet{wang2016learnhist}, although for a purpose of integrating global context in object detection and segmentation tasks.
\citet{khasanova2017graphomni} uses the mean and variance of diffused versions of the last feature maps in their statistical layer.}
\TK{why is this section in TODO?}

@@

% In this work, we are using the graph CNN formulation introduced in~\cite{defferrard2016convolutional}.

@@

A CNN is composed of the following main building blocks~\citep{lecun1998cnn}:
(i) a convolution,
(ii) a non-linearity,
(iii) a down-sampling operation,
(iv) a pooling operation, and
(v), optionally,  normalization.\footnote{\citet{ioffe2015batchnorm} have shown that batch normalization helps training. We verified this experimentally in our setting as well.}
Our architecture is depicted in~\figref{architecture} and discussed in greater details in \secref{architecture}. As operations (ii) and (v) are point-wise, they do not depend on the data domain. The pooling operation is simply a permutation invariant aggregation function which does not need to be adapted either. The convolution and down-sampling operations however need to be generalized from Euclidean domains to the sphere.

On regular Euclidean domains, such as 1-dimensional time series or 2-dimensional images, a convolution can be efficiently implemented by sliding a localized convolution kernel (for example a patch of 5x5 pixels) in the signal domain.
Likewise, down-sampling can be achieved by taking one pixel out of $n$.
Because of the irregular sampling, there is no straightforward way to define a convolution on the sphere directly in the pixel domain; convolutions are most often performed using spherical harmonics.
In our method we also use the spectral domain to define the convolution.
The implementation, however, does not need a direct access to the spectrum, which is computationally more efficient (see \secref{efficient_convolution}).
% We will therefore use the spectral domain to define the convolutions
% \footnote{Note that we only use the spectrum to define the convolution.
% The implementation however does not need a direct access to the spectrum, which is computationally much more efficient. See \secref{efficient_convolution}.}
% \TK{moving footnote to the main text }
The gist of our method is to define the convolution operation on a sphere using a graph, and the down-sampling operation using a hierarchical pixelisation of the sphere.

@@

Before doing any numerical analysis on the sphere, one first has to choose a tessellation, i.e., an exhaustive partition of the sphere into finite area elements, where the data under study is quantized.
The simplicity of the spherical form belies the intricacy of global analysis on the sphere.
There is no known point set which achieves the analogue of uniform sampling in Euclidean space and allows exact and invertible discrete spherical harmonic decompositions of arbitrary but band-limited functions.
Any existing proposition of practical schemes for the discrete treatment of such functions on the sphere introduces some systematic error dependent on the global properties of the point set.
While our method is applicable to any pixelisation of the sphere, two details depend on the chosen sampling: (i) the choice of neighbours in the construction of the graph, and (ii) the choice of parent nodes when coarsening the graph.
As HEALPix~\citep{gorski2005healpix} is the most popular sampling used in cosmology, our target application, we'll tailor the method to that particular sampling in the subsequent exposition.
%We'll note in later sections where the method should be adapted for a different sampling.

@@

HEALPix is an acronym for Hierarchical Equal Area isoLatitude Pixelization of a sphere.
This pixelisation produces a hierarchical subdivision of a spherical surface where each pixel covers the same surface area as every other pixel.
A hierarchy is desired for data locality in computer memory.
Equal area is advantageous because white noise generated by the signal receiver gets integrated exactly into white noise in the pixel space.
Isolatitude is essential for the implementation of a fast spherical transform.
HEALPix is the sole pixelisation scheme which satisfies those three properties.

The lowest possible resolution is given by the base partitioning of the surface into $N_{pix} = 12$ equal sized pixels (right-most sphere in \figref{pooling}).
As each pixel is subdivided in four, the second coarser resolution is $N_{pix} = N_{side}^2 \cdot 12 = 2^2 \cdot 12 = 48$ pixels (middle sphere in \figref{pooling}), the third is $N_{pix} = N_{side}^2 \cdot 12 = 4^2 \cdot 12 = 192$ pixels, etc., where $N_{side} = 1, 2, 4, 8, \ldots$ is the grid resolution parameter.
High-resolutions HEALPix maps easily reach millions of pixels.
\figref{example_maps} shows two examples of HEALPix maps: the Cosmic Microwave Background \citep{planck2015overview} and \TK{update}.

@@

From the HEALPix pixelization, we build a weighted undirected graph $\G = (\V, \E, \W)$, where $\V$ is the set of $N_{pix} = |\V|$ nodes, $\E$ is the set of edges, and $\W$ is the weighted adjacency matrix.
In our graph, each pixel $i$ is represented by a node (also called vertex) $v_i \in \V$.
Each node $v_i$ is then connected to the $8$ (or $7$)\footnote{\label{neighbors}The $12 \cdot 4 = 48$ pixels at the corner of each rhombus of the base dodecahedron only have 7 neighboring pixels. See Figures~\ref{fig:healpix_graph_4} and~\ref{fig:pooling}.} nodes $v_j$ which represent the neighboring pixels $j$ of pixel $i$, forming edges $(v_i, v_j) \in \E$. Given those edges, we define the weighted adjacency matrix $\W \in \R^{N_{pix} \times N_{pix}}$ as
\begin{equation*}
	\W_{ij} = \begin{cases}
		\exp \left( -\frac{\|\x_i-\x_j\|_2^2}{\sigma^2} \right) & \text{if pixels $i$ and $j$ are neighbors,} \\
		0 & \text{otherwise,} \\
	\end{cases}
\end{equation*}
where $\x_i$ is a vector encoding the 3-dimensional coordinates of pixel $i$, and
\begin{equation*}
	\sigma = \frac{1}{|\E|} \sum_{(v_i, v_j) \in \E} \|\x_i-\x_j\|_2
\end{equation*}
is the average Euclidean distance over all connected pixels. This weighting scheme is important as distances between pixels are not equal.
Other weighting schemes are possible. For example,~\cite{khasanova2017graphomni} uses the inverse of the distance instead. We found out that the one proposed above works well for our purpose, and did not investigate other approaches, leaving it to future work.
\figref{healpix_graph_4} shows a graph constructed from the HEALPix sampling of a sphere.

@@

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{eigenvectors}
	\caption{The first 16 eigenvectors of the graph Laplacian, an equivalent of Fourier modes, of a graph constructed from the HEALPix sampling of the sphere.
    Eigenvectors 1--3 could be associated with spherical harmonics of degree $\ell=1$ and order $|m|=(0,1)$, eigenvectors 4--8 with degree $\ell=2$ and order $|m|=(0,1,2)$, and eigenvectors 9--15 with degree $\ell=3$ and order $|m|=(0,1,2,3)$.
    Nevertheless, the graph eigenvectors are only approximating spherical harmonics.}
	\label{fig:graph_harmonics}
\end{figure}

Figure \figref{graph_harmonics} shows the Fourier modes of a HEALPix graph, created using the graph construction described above.
The graph Fourier modes resemble the spherical harmonics.
That is a strong hint that the graph is able to capture the spherical properties of the HEALPix sampling.
This topic is further discussed in \ref{sec:comparison_spherical_harmonics}.

@@

As there is no notion of translation on a graph, we cannot convolve two graph signals in a strict sense.
We can, however, convolve a signal with a kernel defined in the spectral domain.
More precisely, we can filter a graph signal by a kernel.
Given the convolution kernel
$h: \R_+ \rightarrow \R$, a signal $\f \in \R^{N_{pix}}$ on the graph is filtered as
\begin{equation} \label{eqn:graph_convolution}
	h(\L) \f = \U h(\bLambda) \U\trans \f,
\end{equation}
where $h(\bLambda)$ is a diagonal matrix where $h(\bLambda)_{ii} = h(\bLambda_{ii})$, and $h(\bLambda) \hat{\f}$ is the spectral representation of $\f$ after filtering.

Contrary to classical signal processing on Euclidean domains, the kernel $h$ has no single representation in the vertex domain and cannot be translated on the graph. It can however be localized on any vertex $v_i$ by the convolution with a Kronecker delta $\b \delta_i \in \R^{N_{pix}}$,\footnote{A Kronecker delta is the signal $\b \delta_i \in \R^{N_{pix}}$ that is zero everywhere except on vertex $v_i$ where it takes the value one.} that is $h(\L) \b \delta_i$, a column of $h(\L)$.
This localization can be useful to visualize kernels as explained in~\ref{sec:filter_visualization}.
If the graph is not regular, i.e., all nodes do not have the same number of neighbors, and all distances are not equal, the effect of the kernel will slightly differ from one node to another. While there is no perfect sampling of the sphere, these differences are negligible as the structure of the whole graph is very regular. However, when considering only parts of the sphere, one can observe important border effects (see \figref{border_effects} and~\ref{sec:border_effects}).
Finally, the graph convolution can be interpreted in the vertex domain as a scalar product with all localizations of the kernel:
\begin{equation} \label{eqn:graph_convolution_spatial}
	(h(\L) \f)_i = \langle h(\L) \b \delta_i, \f \rangle.
\end{equation}
\textbf{Example.}
Let us consider the heat diffusion problem
\begin{equation} \label{eqn:heat_equation}
	\L \b{f}(t) = \tau \partial_t \b{f}(t),
\end{equation}
where $\b{f}: \R_+ \rightarrow \R^{N_{pix}}$. Given the initial condition
$\b{f}(0)$, the solution of~\eqnref{heat_equation} can be expressed as
\begin{equation*}
	\b{f}(t) = e^{-\L \tau t} \b{f}(0) = \U e^{-\bLambda t \tau} \U\trans \g{f}(0) = K_t(\L) \b{f}(0),
\end{equation*}
which is, by definition, the convolution of the signal $\b{f}(0)$ with the kernel $K_t(\lambda)=e^{-\tau t \lambda}$. Since the kernel $K_t$ is applied to the graph eigenvalues $\bLambda$, which can be interpreted as squared frequencies, it can also be considered as a generalization of the Gaussian kernel on the sphere. \figref{gaussian_filters_comparizon} shows the effect of the convolution by diffusing a unit of heat for $\tau=1$ at various times $t$. By comparing the graph convolution with the spherical symmetric Gaussian smoothing, we observe that both techniques lead to similar results (see \figref{gaussian_filters_comparizon}). While the graph convolution remains different from the spherical convolution, this small experiment shows that, providing the correct parameters, graph convolution can approximate spherical convolution. For additional insides about the graph kernel $K_t$, we refer to the \figref{gaussian_filters_visualization} and~\ref{sec:filter_visualization}.

@@

Coarsening can be naturally designed for hierarchical pixelisation schemes, as each subdivision divides a cell in an equal number of child sub-cells.
Coarsening is the reverse operation: merging the sub-cells toward the goal of summarizing the data supported on them.
Merging cells lead to a coarser graph.
Coarsening defines $\C(i)$, the set of children of node $v_i$.
For the HEALPix subdivision scheme, the number of children is constant, i.e., $| \C(i) | = 4^p \ \forall i$, for some $p$.
